**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**

### 2016-02-12 Schloss_PacBio16S_PeerJ_2016.Rmd
Incorporate Jenior's comments

Historically, all sequencing platforms were created to primarily perform genome sequencing. When sequencing a genome, it is assumed that the same base of DNA will be sequenced multiple times and the consensus of multiple sequence reads is used to generate contigs. Thus, although an individual base call may have a high error rate, the consensus sequence will have a low error rate. To sequence the 16S rRNA gene researchers use conserved primers to amplify a sub-region from within the gene that is isolated from many organisms. Because the fragments are not cloned, it is not possible to obtain high sequence coverage from the same DNA molecule using these platforms. Thus, to reduce sequencing error rates it has become imperative to develop stringent sequence curation and denoising algorithms [@Kozich2013; @Schloss2011a]. There has been a tradeoff between read length, number of reads per sample, and the error rate. For instance, we recently demonstrated that using the Illumina MiSeq and the 454 Titanium platforms the raw error rate varies between 1 and 2% [@Kozich2013; @Schloss2011a]. Yet, it was possible to obtain error rates below 0.02% by adopting various denoising algorithms; however, the resulting fragments were only 250-nt long. In the case of 454 Titanium, extending the length of the fragment introduces length-based errors and in the case of the Illumina MiSeq, increasing the length of the fragment reduces the overlap between the read pairs reducing the ability of each read to mutually reduce the sequencing error. Inadequate denoising of sequencing reads can have many negative effects including limited ability to identify chimeras [@Haas2011; @Edgar2011] and inflation of alpha- and beta-diversity metrics [@Huse2010; @Kunin2010; @Kozich2013; @Schloss2011a]. Illumina's MiSeq plaform enjoys widespread use in the field because of the ability to sequence 15-20 million fragments that can be distributed across hundreds of samples for less than $5000 (US).

@@

Despite the opportunity to potentially generate high-quality full-length sequences, the Pacific Biosciences platform has not been widely adopted for sequencing 16S rRNA genes [@Fichot2013; @Mosher2014; @Mosher2013]. Previous studies utilizing the technology have removed reads with mismatched primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013]. Others have utilized the platform without describing the bioinformatic pipeline that was utilized [@Mosher2014; @Mosher2013]. The only study to report the error rate of the platform for sequencing 16S rRNA genes with a mock community used the P4-C2 chemistry and obtained an error rate of 0.32%, which is 16-fold higher than has been observed using the MiSeq or 454 platforms [@Schloss2011a; @Kozich2013]. In the current study, we assessed the quality of data generated by the PacBio sequencer usingthe improved P6-C4 chemistry and on-sequencer data processing. The goal was to determine whether this strategy could fill the need for generating high-quality, full-length sequence data on par with other platforms. We hypothesized that by modulating the 16S rRNA gene fragment length we could alter the read depth and obtain reads longer than are currently available by the 454 and Illumina platforms but with the same quality. To test this hypothesis, we developed a sequence curation pipeline that was optimized by reducing the sequencing error rate of a mock bacterial community with known composition. The resulting pipeline was then applied to 16S rRNA gene fragments that were isolated from soil and human and mouse feces.

@@

```{r basic}
raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])
```

We established a simple curation procedure by culling any sequence that had a string of the same base repeated more than 8 times  or did not start and end at the expected alignment coordinates for that region of the 16S rRNA gene. This reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=3, nsmall=2)` to `r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%. This basic procedure resulted in the removal of between `r format(min(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(which.min(raw_to_basic_lost))]`) and `r format(max(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(which.max(raw_to_basic_lost))]`)% of the reads (Table 2). Although the percentage of reads removed increased with the length of the fragment, there was no obvious relationship between fragment length and error rate (Figure 2).

@@

Full-length sequences are frequently seen as a panacea to overcome the limitations of taxonomic classifications. The ability to classify each of our sample types benefited from the generation of full-length sequences. It was interesting that the benefit varied by sample type and database. For example, using the mouse libraries, the ability to classify each of the regions differed by less than 5% when classifying against the SILVA and greengenes databases. The effect of the database that was used was also interesting. The RDP database outperformed the other databases for the mouse samples and the SILVA database outperformed the others for the human and soil samples. The three databases were equally effective for classifying the mock community. Finally, since only the greengenes database provided species-level information for its reference sequences it was the only database that allowed for resolution of species-level classification. The sequences from the mouse and soil libraries were not effectively classified to the species level (all less than 10%). In contrast, classification of the human libraries resulted in more than 40% of the sequences being classified to a genus, regardless of the region. That the variation in species-level classification for the human libraries was less than 10% suggests that the benefit of added length is minimal considering the lower sequencing yield.

>>>>

Historically, all sequencing platforms were created to primarily perform genome sequencing. When sequencing a genome, it is assumed that the same base of DNA will be sequenced multiple times and the consensus of multiple sequence reads is used to generate contigs. Thus, an individual base call may have a high error rate, but the consensus sequence will have a low error rate. To sequence the 16S rRNA gene researchers use conserved primers to amplify a sub-region from within the gene that is isolated from many organisms. Because the fragments are not cloned, it is not possible to obtain high sequence coverage from the same DNA molecule using these platforms. To reduce sequencing error rates it has become imperative to develop stringent sequence curation and denoising algorithms [@Kozich2013; @Schloss2011a]. There has been a tradeoff between read length, number of reads per sample, and the error rate. For instance, we recently demonstrated that using the Illumina MiSeq and the 454 Titanium platforms the raw error rate varies between 1 and 2% [@Kozich2013; @Schloss2011a]. Yet, it was possible to obtain error rates below 0.02% by adopting various denoising algorithms; however, the resulting fragments were only 250-nt long. In the case of 454 Titanium, extending the length of the fragment introduces length-based errors and in the case of the Illumina MiSeq, increasing the length of the fragment reduces the overlap between the read pairs reducing the ability of each read to mutually reduce the sequencing error. Inadequate denoising of sequencing reads can have many negative effects including limited ability to identify chimeras [@Haas2011; @Edgar2011] and inflation of alpha- and beta-diversity metrics [@Huse2010; @Kunin2010; @Kozich2013; @Schloss2011a]. Illumina's MiSeq plaform enjoys widespread use in the field because of the ability to sequence 15-20 million fragments that can be distributed across hundreds of samples for less than $5000 (US).

@@

Despite the opportunity to potentially generate high-quality full-length sequences, the Pacific Biosciences platform has not been widely adopted for sequencing 16S rRNA genes [@Fichot2013; @Mosher2014; @Mosher2013]. Previous studies utilizing the technology have removed reads with mismatched primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013]. Others have utilized the platform without describing the bioinformatic pipeline that was utilized [@Mosher2014; @Mosher2013]. The only study to report the error rate of the platform for sequencing 16S rRNA genes with a mock community used the P4-C2 chemistry and obtained an error rate of 0.32%, which is 16-fold higher than has been observed using the MiSeq or 454 platforms [@Schloss2011a; @Kozich2013]. In the current study, we assessed the quality of data generated by the PacBio sequencer using the improved P6-C4 chemistry and on-sequencer data processing. The goal was to determine whether this strategy could fill the need for generating high-quality, full-length sequence data on par with other platforms. We hypothesized that by modulating the 16S rRNA gene fragment length we could alter the read depth and obtain reads longer than are currently available by the 454 and Illumina platforms but with the same quality. To test this hypothesis, we developed a sequence curation pipeline that was optimized by reducing the sequencing error rate of a mock bacterial community with known composition. The resulting pipeline was then applied to 16S rRNA gene fragments that were isolated from soil and human and mouse feces.

@@

```{r basic}
raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])
names(raw_to_basic_lost) <- error_summary[error_summary$method == "basic","region"]
```

We established a simple curation procedure by culling any sequence that had a string of the same base repeated more than 8 times  or did not start and end at the expected alignment coordinates for that region of the 16S rRNA gene. This reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=2, nsmall=2)` to `r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%. This basic procedure resulted in the removal of between `r format(min(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(raw_to_basic_lost[which.min(raw_to_basic_lost)])]`) and `r format(max(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(raw_to_basic_lost[which.max(raw_to_basic_lost)])]`)% of the reads (Table 2). Although the percentage of reads removed increased with the length of the fragment, there was no obvious relationship between fragment length and error rate (Figure 2).

@@

Full-length sequences are frequently seen as a panacea to overcome the limitations of taxonomic classifications. The ability to classify each of our sample types benefited from the generation of full-length sequences. It was interesting that the benefit varied by sample type and database. For example, using the mouse libraries, the ability to classify each of the regions differed by less than 5% when classifying against the SILVA and greengenes databases. The effect of the database that was used was also interesting. The RDP database outperformed the other databases for the mouse samples and the SILVA database outperformed the others for the human and soil samples. The three databases were equally effective for classifying the mock community. Finally, since only the greengenes database provided species-level information for its reference sequences it was the only database that allowed for resolution of species-level classification. The sequences from the mouse and soil libraries were not effectively classified to the species level (all less than 10%). In contrast, classification of the human libraries resulted in more than 40% of the sequences being classified to a genus, regardless of the region. These data demonstrate that for the samples we analyzed, the length of the sequence fragment was not as significant a factor in classification as the choice of database.
